Generalized spaces for constructive algebra

### 2019-09-22 slides-herrsching2019.tex
Slides for Lecture II finished

  \normalsize
\end{frame}

\begin{frame}{Locales presented by theories}
  \justifying
  \textbf{Definition.} A \hil{geometric signature} consists of
  \begin{enumerate}
    \item a set of sorts: $X$, $Y$, $Z$, \ldots
    \item a set of function symbols: $f : X \times Y \to Z$, \ldots
    \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
  \end{enumerate}
  % A geometric signature is \hil{propositional} iff its set
  % of sorts is empty.
  A \hil{geometric theory}~$\TT$ over a geometric
  signature~$\Sigma$ consists of a set of axioms (geometric sequents
  over~$\Sigma$).

  \only<1>{
    \textbf{Examples.}
    \begin{enumerate}
      \item the theory of rings
      \item the theory of surjections~$\NN \twoheadrightarrow \RR$
      \item the theory of Dedekind cuts
    \end{enumerate}
  }
  \pause

  \textbf{Definition.} A \hil{set-based model}~$M$ of a theory~$\TT$ consists of
  \begin{enumerate}
    \item a set~$\brak{X}$ for each sort~$X$,
    \item a function~$\brak{f} : \brak{X_1} \times \cdots \times \brak{X_n} \to
    \brak{Y}$ for \\
    each function symbol~$f : X_1 \times \cdots \times X_n \to
    Y$, and
    \item a relation~$\brak{R} \subseteq \brak{X_1} \times \cdots \times
    \brak{X_n}$ for \\
    each relation symbol~$R \hookrightarrow X_1 \times \cdots \times X_n$
  \end{enumerate}
  such that~$M$ validates the axioms of~$\TT$.
\end{frame}

>>>>

  \begin{hilblock}Sheaves allow us to explore mathematical objects from
  custom-tailored mathematical universes.\phantom{p}\end{hilblock}
\end{frame}

\begin{frame}{Frames presented by theories}
  \justifying
  \textbf{Definition.} A \hil{geometric theory} consists of
  \begin{enumerate}
    \small
    \item a set of sorts: $X$, $Y$, $Z$, \ldots
    \item a set of function symbols: $f : X \times Y \to Z$, \ldots
    \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
    \item a set of axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
  \end{enumerate}
  % A geometric signature is \hil{propositional} iff its set
  % of sorts is empty.

  \only<1->{
    \textbf{Examples.}
    The theory of rings,
    of surjections~\mbox{$\NN \twoheadrightarrow \RR$},
    of Dedekind cuts,
    of prime ideals of a given ring, \ldots
  }

  \textbf{Definition.} A \hil{set-based model}~$M$ of a theory~$\TT$ consists of
  \begin{enumerate}
    \small
    \item a set~$\brak{X}$ for each sort~$X$,
    \item a function~$\brak{f} : \brak{X_1} \times \cdots \times \brak{X_n} \to
    \brak{Y}$ \\
    for each function symbol~$f : X_1 \times \cdots \times X_n \to Y$, and
    \item a relation~$\brak{R} \subseteq \brak{X_1} \times \cdots \times
    \brak{X_n}$ \\
    for each relation symbol~$R \hookrightarrow X_1 \times \cdots \times X_n$
  \end{enumerate}
  such that~$M$ validates the axioms of~$\TT$.

  \jnote{1}{
    A geometric theory is called \emph{propositional} iff its set of
    sorts is empty. Hence the only ingredients of a propositional geometric
    theory are a set of nullary relation symbols (atomic propositions) and a
    set of axioms.

    Associated to any propositional geometric theory~$\TT$ is its
    \emph{Lindenbaum algebra}. This is the set of formulas over the signature
    of the theory modulo provable equivalence. Endowed with the partial order
    given by
    \[ [\varphi] \leq [\psi] \quad\vcentcolon\Longleftrightarrow\quad
      \text{$\TT$ proves $\varphi \vdash \psi$}, \]
    the Lindenbaum algebra becomes a frame.

    The Lindenbaum algebra of~$\TT$ is the free frame generated by the
    atomic propositions of~$\TT$ modulo the axioms of~$\TT$. The associated
    locale~$L(\TT)$ is the \emph{classifying locale of~$\TT$}. Its points are
    in canonical bijection with the set-based models of~$\TT$.
  }
\end{frame}

\begin{frame}{Sheaves}
  \justifying
  \textbf{Definition.} A \hil{presheaf}~$F$ on a locale~$X$ is given by
  \begin{enumerate}
    \item a set~$F(U)$ for each open~$U$ of~$X$ and
    \item a map~$(\cdot)|^U_V : F(U) \to F(V)$ for each opens~$V \leq U$
  \end{enumerate}
  such that $(\cdot)|^U_U = \operatorname{id}_{F(U)}$ for all~$U$ and
  $(\cdot)|^V_W \circ (\cdot)|^U_V = (\cdot)|^U_W$ for all~$W \leq V
  \leq U$.
  A \hil{compatible family} with respect to a covering~$U = \bigvee_i U_i$
  is a family~$(s_i)_i$ of sections~$s_i \in F(U_i)$ such that~$s_i|^{U_i}_{U_i
  \wedge U_j} = s_j|^{U_j}_{U_i \wedge U_j}$. $F$ is a \hil{sheaf} iff for any
  such family there is a unique section~$s \in F(U)$ such that~$s|^U_{U_i} =
  s_i$ for all~$i$.

  \textbf{Example.} The \hil{sheaf of continuous functions} on a locale~$X$ is
  given by~$\C(U) = \Hom(U,\RR)$.

  \textbf{Non-example.} The \hil{presheaf of constant
  functions},~$\C_{\text{c}}(U) = \text{``the set of constant functions~$U \to
  \RR$''}$, it usually not a sheaf.
\end{frame}

\begin{frame}{Sheaf semantics}
  \justifying\small
  Let~$X$ be a locale. We recursively define what it means for a
  \hil{first-order formula over an open~$U$ of~$X$} to be forced:
  \[ \footnotesize\renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \bot \\
  U \models s = t &\Ll& \text{$s = t$ when evaluated as elements of~$F(U)$} \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models
  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigvee_i U_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all~$V \leq U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all $V \leq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $U = \bigvee_i U_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
  \end{array} \]
  \textbf{Theorem.} If~$U \models \varphi$ and if~$\varphi$ entails~$\psi$
  intuitionistically, then~$U \models \psi$.

  \jnote{1}{
    A \emph{formula over an open~$U$} of~$X$ is a first-order formula (made up
    using~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\ {\Rightarrow}\ {\forall}\
    {\exists}$) over the signature which has one sort for each sheaf~$F$, one
    constant symbol of sort~$F$ for each section~$s \in F(U)$, one function
    symbol~$f : F \to G$ for each morphism of sheaves, and so on.
  }
\end{frame}

\begin{frame}{Internalizing parameter-dependence}
  \justifying
  \small
  Let~$X$ be a topological space. Let~$(f_x)_{x \in X}$ be a continuous family of continuous
  functions~$\RR \to \RR$ (that is, let a continuous function $X \times \RR \to \RR,\,(x,a) \mapsto
  f_x(a)$ be given). This family defines a sheaf morphism~$f : \C \to
  \C$ by~$f_U : \C(U) \to \C(U),\,s \mapsto (x \mapsto f_x(s(x)))$.
  \normalsize

  \begin{itemize}\justifying
    \item $X \models \text{``$\C$ is the set of Dedekind real numbers''}.$
    \item $X \models \text{``the function $f : \C \to \C$ is
    continuous''}$.\medskip
    \item Iff $f_x(-1) < 0$ for all $x \in X$, then $X \models f(-1) < 0$.
    \item Iff $f_x(+1) > 0$ for all $x \in X$, then $X \models f(+1) > 0$.
    \item Iff all $f_x$ are increasing, then $X \models \text{``$f$ is
    increasing''}$.\medskip
    \item Iff there is an open cover~$X = \bigcup_i U_i$ such
    that for each~$i$, there is a continuous function~$s : U_i \to \RR$
    with~$f_x(s(x)) = 0$ for all~$x \in U_i$, then $X \models \exists s
    \? \C\_ f(s) = 0$.
  \end{itemize}

  \jnote{1}{Hence:
  \begin{enumerate}\justifying
    \item The standard formulation of the intermediate value theorem fails
    in~$\Sh(X)$, because its external interpretation is that in continuous
    families of continuous functions, zeros can locally be picked continuously.
    That claim is false, as
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
    counterexample} demonstrates.

    As a corollary, we deduce that the standard formulation of the
    intermediate value theorem is not constructively provable.

    \item The \emph{approximate} intermediate value theorem (stating
    that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
    \varepsilon$)
    \fixedhref{https://mathoverflow.net/questions/253059/approximate-intermediate-value-theorem-in-pure-constructive-mathematics}{has
    a constructive proof} and therefore holds in~$\Sh(X)$. The
    external interpretation is that in continuous families of continuous
    functions, approximate zeros can locally be picked continuously.

    \item The \emph{monotone} intermediate value theorem, stating that a strictly
    increasing continuous function with opposite signs has a unique zero,
    admits a constructive proof and therefore holds in~$\Sh(X)$. The external
    interpretation is that in continuous families of strictly increasing
    continuous functions, zeros can globally be picked continuously. You are
    invited to prove this fact directly, without reference to the internal
    language. This exercise isn't particularly hard, but it's not trivial either.
  \end{enumerate}}
\end{frame}
